\normalfont\spacedallcaps{Hazardous asteroids forecast via Markov random fields}

### 2021-11-28 Manuscript/manuscript.tex
PUB

\section*{Abstract} The machine learning algorithms provide a promising approach to classify and predict natural or social phenomena. Differently to the oldest theoretical approach in which the laws that describe particular phenomena are derived from general assumptions using a mathematical language (e.g., the principle of energy conservation, the three Newton laws, the second principle of thermodynamics \cite{feynman2018feynman}), the machine learning algorithms start from very general assumptions/expressions and then their parameters, are optimized with the likehood maximization\cite{russell2010artificial,murphy2012machine}. The first approach allows obtaining interpretable predictions, the second one good forecast with a reduced effort. However, the tool that has to be paid in this case is the low interpretability \cite{russell2010artificial,murphy2012machine}. Since they provide the list of conditional independences/dependencies, the graphical methods represent a good compromise between the need for interpretability and the forecasts obtained without developing a general theory. In order to prove the validity of this statement, I considered a dataset for which the theory that interconnects its feature was known: in particular, I chose the asteroids hazardousness as provided by CNEOS \cite{cneos+nasa} and published on Kaggle \cite{kaggle_dataset}. The outcomes of the probabilistic methods for predicting the asteroid hazardousness were compared with the ones provided by the theory and with the ones of Random Forest, Support Vector Machines, and Logistic Regression and Quadric Discriminant analysis. The results show that the forecast performances of probabilistic methods are better than QDA and almost equal to the logistic regression but lower to RF and SVM. However, since the list of conditional dependencies correctly reflects the laws of celestial mechanics \cite{murray1999solar}, it can be said that in this case, the probabilistic methods provide an interpretable and correct explanation of their mechanism. Therefore, contrary to other machine learning algorithms, such methods can be fully validated scientifically.

@@

\chapter{Introduction}

The description and the forecast of physical \footnote{Where here physical should be inteded as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms one can formulate theoretical models that provide equations that describe such phenomena. Such approach, which will be called here \textit{ab-initio}. On the other side one can start from data and fit them into some general models (with fixed or not fixed number of parameters). This method will be called here as machine learning (ML) method. In the first case one have the advantage to have an almost full explanation of the meaning of the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated with respect to the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case for instance of Hartree-Fock method or of Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation in to a much more computationally affordable way. However also in this case the solution for solids can be very expensive \cite{martin_2004}} . On the other side, the ML methods since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require a development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The toll that has to be paid in this case it that the underlying mechanism by which a forecast is preferred to another one may be not so interpretable. Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century astronomers know that mercury follows an elliptical orbit that instead of being fixed rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newtons gravity law) or instead reformulate from scratch the Newton gravity. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise  between these two approaches: they can be used for every physical phenomena and, since they provide the conditional dependencies/independences they also give a fine and intuitive representation  of their mechanisms. Therefore, if the theory of the process is known, one can use this class of algorithm and compare its finding with the theory. In this way one can not only evaluate the quality of the forecasts, but also of the model provided by the graphical methods. The aim of the present work is to perform such analysis for a dataset that contain the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. The present work is organized as follows: first a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described and finally, the main results here obtained with graphical methods will be presented as compared with other type of machine learning algorithms. Furthermore an appendix about the theoretical concepts of the theory that explains and interconnects the dataset features is also provided at the end of the present work.

>>>>

\section*{Abstract} The machine learning algorithms provide a promising approach to classify and predict natural or social phenomena. Differently to the oldest theoretical approach in which the laws that describe particular phenomena are derived from general assumptions using a mathematical language (e.g., the principle of energy conservation, the three Newton laws, the second principle of thermodynamics \cite{feynmanlectures}), the machine learning algorithms start from very general assumptions/expressions and then their parameters, are optimized with the likehood maximization\cite{russell2010artificial,murphy2012machine}. The first approach allows obtaining interpretable predictions, the second one good forecast with a reduced effort. However, the toll that has to be paid in this case is the low interpretability \cite{russell2010artificial,murphy2012machine}. Since they provide the list of conditional independences/dependencies, the graphical methods represent a good compromise between the need for interpretability and the forecasts obtained without developing a general theory. In order to prove the validity of this statement, I considered a dataset for which the theory that interconnects its feature was known: in particular, I chose the asteroids hazardousness as provided by CNEOS \cite{cneos+nasa} and published on Kaggle \cite{kaggle_dataset}. The outcomes of the probabilistic methods for predicting the asteroid hazardousness were compared with the ones provided by the theory and with the ones of Random Forest (RF), Support Vector Machines (SVM), and Logistic Regression and Quadric Discriminant analysis (QDA). The results show that the forecast performances of probabilistic methods are better than QDA and almost equal to the logistic regression but lower to RF and SVM. However, since the list of conditional dependencies correctly reflects the laws of celestial mechanics \cite{murray1999solar}, it can be said that in this case, the probabilistic methods provide an interpretable and correct explanation of their mechanism. Therefore, contrary to other machine learning algorithms, such methods can be fully validated scientifically.

@@

\chapter{Introduction} The description and the forecast of physical \footnote{Where here physical should be intended as mesuarable} phenomena can be attacked with two different approaches: starting from a restricted set of principles/axioms, one can formulate theoretical models that provide equations that describe such phenomena. Such approach, which will be called here \textit{ab-initio}. On the other side, one can start from data and fit them into some general models (with fixed or not fixed number of parameters). These methods will be called here the machine learning (ML) methods. In the first case, one can have a full explanation of the laws obtained and an overall idea of why nature works in this way. The toll that has to be paid in this case is that, since the phenomena are much more complicated than the starting principles, a consistent part of them may be excluded by the assumption made at the beginning. Furthermore, the calculation of the solution given the dynamical equation can be computationally expensive\footnote{This is the case, for instance, of the Hartree-Fock method or Density Functional Theory: such methods rewrite, with some approximations, the Schroedinger equation into a much more computationally affordable way. However, also, in this case, the solution for solids can be costly \cite{martin_2004}}. On the other side, the ML methods, since they make much more general assumptions and work directly on data, are not limited to a particular class of phenomena. Therefore they do not require the development of a specific theory as the \textit{ab-initio} one: the same model can be applied to exoplanets habitability as well to credit risk evaluation. The drawback of this approach is that the underlying mechanism by which a forecast is preferred to another one may not be interpretable.
Such dichotomy between these two methods can be explained with an example taken from astronomy: at the beginning of XX century, astronomers knew that Mercury follows an elliptical orbit that, instead of being fixed, rotates around the Sun as shown in Fig. \ref{Perihelion_precession2} Such phenomenon is called \textit{Precession of the perihelion of Mercury}. The crucial point is that there was no way to explain from the Newton gravity theory such phenomenon. At this point, scholars had two paths: describe its motion with an empirical law (perhaps with an empirical modification of Newton's gravity law) or reformulate the Newton gravity from scratch. The second approach was the one followed by A.Einstein with his general relativity theory (further details are given in the A.Zee textbook \cite{zee2013einstein}). The graphical methods represent a good compromise between these two approaches: they can be used for every physical phenomenon, and, since they provide conditional dependencies/independences, they also give an elegant and intuitive representation of their mechanisms. Therefore, if the process theory is known, one can use this class of algorithms and compare its finding with the theory. In this way, one can evaluate the quality of the forecasts and the model provided by the graphical methods. The present work aims to perform such analysis for a dataset that contain the features and the hazardousness \footnote{For Earth} of asteroids. For this dataset, indeed, the relationships between the various features are known. The present work is organized as follows: first, a brief introduction about the basic theoretical concepts of the statistical methods here used will be provided, then the dataset will be described, and finally, the principal results here obtained with graphical methods will be presented as compared with other types of machine learning algorithms. Furthermore, a recap of the celestial mechanics will be provided in Appendix A. This theory explains the connections and the meaning of the different features contained in the dataset.
